---
title: "State Park Attendance Analytics for Resource Optimization"
author: "Prince Peter Yalley"
date: "2025-01-27"
output: html_document
---

1. Load Libraries
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(readr)
library(scales)
library(forecast)
library(caret)
library(plotly)
```

2. Load Dataset
```{r}
data <- read_csv("State_Park_Annual_Attendance_Figures_by_Facility___Beginning_2003_20250211.csv")
```

3. View Data
```{r}
head(data)
```

4. Data Structure
```{r}
str(data)
```
5. Clean Column Names
```{r}
colnames(data) <- make.names(colnames(data))
```

6. Handle Missing Values
```{r}
data <- na.omit(data)
```

7. Summary Statistics
```{r}
summary(data)
```

8. Convert Year to Numeric
```{r}
data$Year <- as.numeric(data$Year)
```

9. Total Attendance by Year
```{r}
yearly_attendance <- data %>%
  group_by(Year) %>%
  summarise(Total = sum(Attendance))
```

10. Plot Attendance Trend
```{r}
ggplot(yearly_attendance, aes(x = Year, y = Total)) +
  geom_line() +
  geom_point() +
  ggtitle("Annual Park Attendance Trend")
```
```{r}
colnames(data) <- make.names(colnames(data))

```

```{r}
colnames(data)
```
```{r}
top_facilities <- data %>%
  group_by(Facility) %>%
  summarise(Total = sum(Attendance)) %>%
  arrange(desc(Total)) %>%
  head(10)
```

```{r}
top_facilities <- data %>%
  group_by(Facility) %>%
  summarise(
    Total_Attendance = sum(Attendance),
    Avg_Attendance = mean(Attendance)
  ) %>%
  arrange(desc(Total_Attendance)) %>%
  slice_head(n = 10)
```

```{r}
ggplot(top_facilities, aes(x = reorder(Facility, Total_Attendance), y = Total_Attendance)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 State Park Facilities by Attendance",
    x = "Facility",
    y = "Total Attendance"
  )
```

11. Top Facilities
```{r}
top_facilities <- data %>%
  group_by(Facility) %>%
  summarise(
    Total = sum(Attendance),
    Avg_Attendance = mean(Attendance)
  ) %>%
  arrange(desc(Total)) %>%
  slice_head(n = 10)
```

12. Plot Top Facilities
```{r}
ggplot(top_facilities, aes(x = reorder(Facility, Total), y = Total)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Facilities by Attendance",
    x = "Facility",
    y = "Total Attendance"
  )
```

13. Year-over-Year Growth
```{r}
yearly_attendance <- data %>%
  group_by(Year) %>%
  summarise(Total = sum(Attendance)) %>%
  arrange(Year)

yearly_attendance <- yearly_attendance %>%
  mutate(Growth = (Total - lag(Total)) / lag(Total))
```

14. Plot Growth
```{r}
ggplot(yearly_attendance, aes(x = Year, y = Growth)) +
  geom_line() +
  geom_point() +
  labs(title = "Year-over-Year Growth Rate")
```

15. Time Series Conversion
```{r}
library(forecast)
```

```{r}
ts_data <- ts(yearly_attendance$Total, start = min(yearly_attendance$Year))
```


```{r}
model <- auto.arima(ts_data)
forecasted <- forecast(model, h = 5)
plot(forecasted)
```

18. Train-Test Split

```{r}
library(caret)
```


```{r}
trainIndex <- createDataPartition(yearly_attendance$Total, p = 0.8, list = FALSE)

train <- yearly_attendance[trainIndex, ]
test <- yearly_attendance[-trainIndex, ]
```

19. Linear Regression Model
```{r}
lm_model <- lm(Total ~ Year, data = train)

predictions <- predict(lm_model, newdata = test)

RMSE(predictions, test$Total)
```

20. Predictions
```{r}
library(plotly)
```

21. Model Evaluation
```{r}
RMSE(predictions, test$Total)
```

22. Interactive Plot - Smooth Line
```{r}
p <- ggplot(yearly_attendance, aes(x = Year, y = Total)) + geom_line()
ggplotly(p)
```

Interactive Plot - Smooth Line

```{r}
p <- ggplot(yearly_attendance, aes(x = Year, y = Total)) +
  geom_line() +
  geom_point()

ggplotly(p)
```

23. Insights
```{r}
print("Attendance trends show long-term growth with fluctuations tied to external factors.")
```

24. Business Recommendation
```{r}
print("Invest in high-performing parks and optimize low-performing facilities using predictive insights.")
```

25. Export Results
```{r}
write.csv(yearly_attendance, "processed_attendance.csv", row.names = FALSE)
```
