---
title: "New York State Licensed Guides"
subtitle: "An Analysis of Outdoor Recreation Leadership Across the Empire State"
author: "Prince Peter Yalley"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    css: report-style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%"
)
```

```{r libraries}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(knitr)
library(kableExtra)
library(lubridate)
library(stringr)
library(forcats)
```

---

# About This Dataset

New York State requires professional guides to obtain a license through the Department of Environmental Conservation (DEC). These individuals lead clients through the state's vast natural landscapes — from the Adirondack High Peaks to the whitewater rivers of the Catskills — for activities including hiking, fishing, hunting, camping, and technical climbing.

This dataset, published by NY Open Data and last updated **January 27, 2025**, captures a snapshot of all **currently licensed guides** in the state. It contains `r nrow(read_csv("data/guides-currently-licensed-in-new-york-state-20250127.csv"))` license records — but because guides can hold multiple activity certifications, the actual number of individual guides is smaller.

**Key questions this analysis explores:**

- Where are guides concentrated geographically in New York?
- What outdoor activities are most represented?
- How does license expiration timing break down?
- What share of licensed guides come from outside New York State?

---

```{r load-data}
raw_data <- read_csv("guides-currently-licensed-in-new-york-state-20250127.csv",
                     show_col_types = FALSE)

# Clean data
guides <- raw_data %>%
  filter(!is.na(County), str_trim(County) != "") %>%
  mutate(
    County = str_to_title(str_trim(County)),
    City   = str_to_title(str_trim(City)),
    `Activity Type Description` = str_trim(`Activity Type Description`),
    Expiration_Date = mdy(`Expiration Date`),
    Exp_Year = year(Expiration_Date),
    NY_Resident = ifelse(State == "NY", "New York", "Out of State")
  )
```

# Overview: The Numbers at a Glance

```{r summary-stats}
n_records   <- nrow(guides)
n_guides    <- n_distinct(guides$`Badge Number`)
n_counties  <- n_distinct(guides$County)
n_activities <- n_distinct(guides$`Activity Type Description`)
n_outofstate <- sum(guides$State != "NY")
pct_outofstate <- round(100 * n_outofstate / n_records, 1)
```

<div class="stat-grid">
<div class="stat-card">
  <div class="stat-num">`r format(n_records, big.mark=",")`</div>
  <div class="stat-label">Total License Records</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r format(n_guides, big.mark=",")`</div>
  <div class="stat-label">Individual Guides</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r n_counties`</div>
  <div class="stat-label">Counties Represented</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r n_activities`</div>
  <div class="stat-label">Activity Types</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r pct_outofstate`%</div>
  <div class="stat-label">Out-of-State Guides</div>
</div>
</div>

---

# Geographic Distribution

## Top 15 Counties by Number of Licensed Guides

New York's outdoor guiding industry is heavily concentrated in the **Adirondack region** — a 6-million-acre park in the northeastern part of the state. Essex, Warren, and Franklin counties — all Adirondack counties — dominate the rankings. The Catskills (Ulster) and the Finger Lakes / Great Lakes corridor also show strong representation.

```{r county-chart, fig.height=7}
county_summary <- guides %>%
  group_by(County) %>%
  summarise(Num_Licenses = n(), .groups = "drop") %>%
  arrange(desc(Num_Licenses)) %>%
  slice_head(n = 15)

ggplot(county_summary, aes(x = fct_reorder(County, Num_Licenses), y = Num_Licenses)) +
  geom_col(fill = "#2c7bb6", width = 0.72) +
  geom_text(aes(label = Num_Licenses), hjust = -0.15, size = 3.4, color = "#333333") +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Top 15 New York Counties by Licensed Guide Count",
    subtitle = "Adirondack counties dominate — Essex alone accounts for 6.5% of all records",
    x = NULL,
    y = "Number of License Records",
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption  = element_text(color = "#999999", size = 9),
    axis.text.y   = element_text(size = 11)
  )
```

## County-Level Data Table

```{r county-table}
guides %>%
  group_by(County) %>%
  summarise(
    `License Records` = n(),
    `Unique Guides`   = n_distinct(`Badge Number`),
    `Avg Licenses/Guide` = round(n() / n_distinct(`Badge Number`), 1)
  ) %>%
  arrange(desc(`License Records`)) %>%
  head(20) %>%
  kable(caption = "Top 20 Counties — License records, unique guides, and certification depth") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left") %>%
  row_spec(0, bold = TRUE, background = "#2c7bb6", color = "white")
```

---

# Activity Types

## What Are Licensed Guides Certified For?

```{r activity-chart, fig.height=6}
activity_summary <- guides %>%
  group_by(`Activity Type Description`) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))

activity_colors <- c(
  "Boats and Canoes" = "#1a7abf",
  "Fishing"          = "#2eaa62",
  "Hiking"           = "#e87722",
  "Camping"          = "#9b59b6",
  "Hunting"          = "#c0392b",
  "WW Rafting"       = "#16a085",
  "Tier I Rock"      = "#8e8e8e",
  "Tier I Ice"       = "#5dade2",
  "Tier II Rock"     = "#626567",
  "Tier II Ice"      = "#85c1e9",
  "WW Kayaking"      = "#48c9b0",
  "WW Canoeing"      = "#58d68d"
)

ggplot(activity_summary,
       aes(x = fct_reorder(`Activity Type Description`, Count),
           y = Count,
           fill = `Activity Type Description`)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_text(aes(label = Count), hjust = -0.15, size = 3.5) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = activity_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Number of License Records by Activity Type",
    subtitle = "Boats & Canoes, Fishing, and Hiking account for nearly two-thirds of all certifications",
    x = NULL,
    y = "Number of License Records",
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption  = element_text(color = "#999999", size = 9),
    axis.text.y   = element_text(size = 11)
  )
```

## Activity Breakdown by Top Counties

This heatmap shows how activity preferences vary across the 10 most guide-rich counties. Counties like **Essex** (Adirondacks) have a diverse spread, while others specialize heavily in water or fishing activities.

```{r heatmap, fig.height=6}
top10_counties <- county_summary$County[1:10]

heatmap_data <- guides %>%
  filter(County %in% top10_counties) %>%
  group_by(County, `Activity Type Description`) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(County) %>%
  mutate(Pct = Count / sum(Count) * 100)

ggplot(heatmap_data,
       aes(x = `Activity Type Description`,
           y = fct_reorder(County, -Count, sum),
           fill = Pct)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(Pct >= 5, paste0(round(Pct), "%"), "")),
            size = 3, color = "white") +
  scale_fill_gradient(low = "#d6eaf8", high = "#1a5276",
                      name = "% of County\nCertifications") +
  labs(
    title    = "Activity Mix Across Top 10 Counties (% of county's certifications)",
    subtitle = "Darker = higher share. Adirondack counties show diverse activity profiles.",
    x = NULL, y = NULL,
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#555555", size = 10),
    axis.text.x   = element_text(angle = 35, hjust = 1, size = 10),
    axis.text.y   = element_text(size = 11),
    legend.position = "right",
    plot.caption  = element_text(color = "#999999", size = 9)
  )
```

---

# License Expiration Timeline

## When Do Current Licenses Expire?

Understanding the expiration wave is important for workforce planning. NY guide licenses appear to be issued on a rolling multi-year basis. The chart below shows how many license records expire in each upcoming year.

```{r expiration-chart}
exp_summary <- guides %>%
  filter(!is.na(Exp_Year)) %>%
  group_by(Exp_Year) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(exp_summary, aes(x = factor(Exp_Year), y = Count)) +
  geom_col(fill = "#e67e22", width = 0.65) +
  geom_text(aes(label = format(Count, big.mark = ",")), vjust = -0.4, size = 3.6) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                     labels = comma) +
  labs(
    title    = "License Records by Expiration Year",
    subtitle = "2028–2029 represent the largest renewal wave; ~13% expire in 2025",
    x = "Expiration Year",
    y = "Number of License Records",
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption  = element_text(color = "#999999", size = 9)
  )
```

---

# Where Do Guides Come From?

## New York Residents vs. Out-of-State Guides

A significant share of licensed guides — roughly **`r pct_outofstate`%** — reside outside New York State. This reflects the regional nature of outdoor guiding: guides from neighboring states like Pennsylvania, New Jersey, Vermont, and Connecticut frequently work in NY's border wilderness areas.

```{r residency-chart, fig.height=5}
state_counts <- guides %>%
  count(State, sort = TRUE) %>%
  filter(State != "NY") %>%
  slice_head(n = 10) %>%
  mutate(State = fct_reorder(State, n))

ggplot(state_counts, aes(x = State, y = n)) +
  geom_col(fill = "#27ae60", width = 0.68) +
  geom_text(aes(label = n), hjust = -0.15, size = 3.5) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Top 10 Out-of-State Home States for NY-Licensed Guides",
    subtitle = "Pennsylvania and New Jersey neighbors contribute the most out-of-state guides",
    x = NULL, y = "Number of License Records",
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption  = element_text(color = "#999999", size = 9)
  )
```

---

# Multi-Certification Guides

## How Diverse Are Guides' Certifications?

Many licensed guides hold certifications in more than one activity — a sign of professional versatility. The chart below shows how certifications are distributed among individual guides (by badge number).

```{r multi-cert}
badge_certs <- guides %>%
  group_by(`Badge Number`) %>%
  summarise(Num_Certs = n(), .groups = "drop") %>%
  count(Num_Certs, name = "Num_Guides")

ggplot(badge_certs, aes(x = factor(Num_Certs), y = Num_Guides)) +
  geom_col(fill = "#8e44ad", width = 0.65) +
  geom_text(aes(label = Num_Guides), vjust = -0.4, size = 3.6) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Distribution of Certifications per Individual Guide",
    subtitle = "Over 1,900 guides hold 2+ certifications — a marker of professional breadth",
    x = "Number of Activity Certifications",
    y = "Number of Guides",
    caption = "Source: NY Open Data | January 2025"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "#555555", size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption  = element_text(color = "#999999", size = 9)
  )
```

---

# Key Takeaways

<div class="takeaway-box">

1. **The Adirondacks are the beating heart of NY guiding.** Essex, Warren, and Franklin counties alone represent over 16% of all license records — a direct reflection of the Adirondack Park's dominance in outdoor recreation.

2. **Water activities lead the way.** Boats & Canoes and Fishing together account for nearly half of all certifications, underscoring New York's exceptional waterway and lake systems.

3. **A regional workforce, not just a local one.** Nearly 1 in 6 licensed guides lives outside New York, with strong representation from PA, NJ, VT, and CT — states that share wilderness borders with NY.

4. **A big renewal wave is coming.** Over 2,900 license records (nearly half) expire in 2028–2029. The NY DEC should anticipate a major renewal processing period in those years.

5. **Professional guides are versatile.** The majority of individual guides (by badge number) hold certifications in 2 or more activity types, reflecting a serious, multi-disciplinary outdoor profession.

</div>

---

*Data sourced from [NY Open Data](https://data.ny.gov/). Analysis conducted in R using `readr`, `dplyr`, `ggplot2`, `lubridate`, and `kableExtra`.*
