---
title: "Labor Market Unemployment Trends Analysis"
subtitle: "Employment, Unemployment & Labor Force Dynamics — 2007 to 2019"
author: "Prince Peter Yalley"
date: "20th May 2025"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
    css: report_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 10,
  fig.height= 6,
  fig.align = "center",
  out.width = "100%"
)
```

```{r libraries}
library(readxl)
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)
library(tidyr)
library(knitr)
library(kableExtra)
library(zoo)
library(forecast)
library(tseries)
```

---

# About This Dataset

This dataset captures **monthly labor market indicators** from **January 2007 through August 2019** — a pivotal 12-year stretch in economic history that includes:

- The **2008 Global Financial Crisis** — the worst economic downturn since the Great Depression
- The **slow but sustained recovery** from 2010 through 2014
- The **post-recession expansion** from 2015 to 2019

Each record represents one month and includes the civilian population, labor force size, participation rate, number of employed, number of unemployed, and both the employment and unemployment rates.

**Key questions this analysis answers:**

- How did the Financial Crisis reshape the labor market, and how long did recovery take?
- Is there a seasonal pattern to unemployment each year?
- What is the relationship between labor force participation and unemployment?
- Based on historical trends, where was unemployment headed after 2019?

---

```{r load-data}
raw <- read_excel("../data/raw/Employment__Unemployment__and_Labor_Force_Data.xlsx")

df <- raw %>%
  mutate(
    Date  = as.Date(Date),
    Year  = as.integer(Year),
    Month = as.integer(Month),
    MonthLabel = month.abb[Month],
    Era = case_when(
      Year <= 2008 ~ "Pre-Crisis",
      Year <= 2010 ~ "Crisis",
      Year <= 2013 ~ "Early Recovery",
      Year <= 2016 ~ "Late Recovery",
      TRUE         ~ "Expansion"
    ),
    Era = factor(Era, levels = c("Pre-Crisis","Crisis","Early Recovery","Late Recovery","Expansion"))
  )
```

# Overview: The Numbers at a Glance

```{r kpis}
peak_unrate  <- df %>% slice_max(`Unemployment Rate`, n=1)
low_unrate   <- df %>% slice_min(`Unemployment Rate`, n=1)
avg_unrate   <- round(mean(df$`Unemployment Rate`), 2)
total_months <- nrow(df)
max_unemp    <- max(df$Unemployed)
min_unemp    <- min(df$Unemployed)
```

<div class="stat-grid">
<div class="stat-card">
  <div class="stat-num">`r total_months`</div>
  <div class="stat-label">Months of Data</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r avg_unrate`%</div>
  <div class="stat-label">Avg Unemployment Rate</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r peak_unrate$'Unemployment Rate'`%</div>
  <div class="stat-label">Peak Rate (Jan 2010)</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r low_unrate$'Unemployment Rate'`%</div>
  <div class="stat-label">Lowest Rate (Dec 2007)</div>
</div>
<div class="stat-card">
  <div class="stat-num">`r format(max_unemp, big.mark=",")`</div>
  <div class="stat-label">Peak Unemployed (persons)</div>
</div>
</div>

---

# Unemployment Rate Over Time

## The Full 2007–2019 Journey

This chart tells the most important story in the data. Starting from a stable pre-recession rate near 3.5%, unemployment exploded during the 2008–2009 Financial Crisis, peaking at **7.8% in early 2010**. What followed was one of the longest labor market recoveries in modern history — it took nearly **7 years** to return to pre-crisis levels.

```{r unemp-timeline}
era_bands <- data.frame(
  xmin  = as.Date(c("2007-01-01","2008-09-01","2010-01-01","2013-01-01","2016-01-01")),
  xmax  = as.Date(c("2008-08-31","2009-12-31","2012-12-31","2015-12-31","2019-08-01")),
  label = c("Pre-Crisis","Crisis","Early Recovery","Late Recovery","Expansion"),
  fill  = c("#e8f5e9","#ffebee","#fff3e0","#e3f2fd","#f3e5f5")
)

ggplot(df, aes(x = Date, y = `Unemployment Rate`)) +
  geom_rect(data = era_bands,
            aes(xmin=xmin, xmax=xmax, ymin=-Inf, ymax=Inf, fill=label),
            inherit.aes = FALSE, alpha = 0.45) +
  scale_fill_manual(values = setNames(era_bands$fill, era_bands$label), name = "Era") +
  geom_line(color = "#c0392b", linewidth = 1.1) +
  geom_point(data = filter(df, `Unemployment Rate` == max(`Unemployment Rate`)),
             color = "#c0392b", size = 4, shape = 21, fill = "white", stroke = 2) +
  geom_point(data = filter(df, `Unemployment Rate` == min(`Unemployment Rate`)),
             color = "#27ae60", size = 4, shape = 21, fill = "white", stroke = 2) +
  annotate("text", x = as.Date("2010-06-01"), y = 8.0,
           label = "Peak: 7.8%\n(Jan 2010)", color = "#c0392b", size = 3.5, fontface="bold") +
  annotate("text", x = as.Date("2007-06-01"), y = 3.0,
           label = "Low: 3.3%\n(Dec 2007)", color = "#27ae60", size = 3.5, fontface="bold") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Monthly Unemployment Rate — 2007 to 2019",
    subtitle = "The 2008 Financial Crisis drove unemployment from 3.5% to 7.8%; full recovery took ~7 years",
    x = NULL, y = "Unemployment Rate (%)",
    caption = "Source: Employment, Unemployment & Labor Force Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

## Yearly Average Unemployment Rate

```{r yearly-avg}
yearly <- df %>%
  group_by(Year) %>%
  summarise(
    Avg_UnempRate  = round(mean(`Unemployment Rate`), 2),
    Avg_LFPRate    = round(mean(`Labor Force Participation Rate`), 2),
    Avg_Employed   = round(mean(Employed)),
    Avg_Unemployed = round(mean(Unemployed)),
    .groups = "drop"
  )

ggplot(yearly, aes(x = factor(Year), y = Avg_UnempRate,
                   fill = Avg_UnempRate)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = paste0(Avg_UnempRate, "%")), vjust = -0.4, size = 3.6) +
  scale_fill_gradient2(low = "#27ae60", mid = "#f39c12", high = "#c0392b",
                       midpoint = 5.5, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Average Annual Unemployment Rate",
    subtitle = "Color intensity reflects severity — green=low, red=high",
    x = NULL, y = "Avg Unemployment Rate (%)",
    caption = "Source: Employment, Unemployment & Labor Force Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

---

# Labor Force Participation Rate

## Participation vs. Unemployment — A Diverging Story

One of the most important — and often overlooked — labor market signals is the **Labor Force Participation Rate (LFPR)**. It measures the share of the working-age population that is either employed or actively looking for work.

What makes this dataset fascinating is that **LFPR declined steadily even as unemployment fell**. This means some of the unemployment "improvement" came not from people finding jobs, but from people *giving up looking* — dropping out of the labor force entirely.

```{r lfpr-vs-unemp}
df_long <- df %>%
  select(Date, `Unemployment Rate`, `Labor Force Participation Rate`) %>%
  pivot_longer(-Date, names_to = "Metric", values_to = "Value")

ggplot(df_long, aes(x = Date, y = Value, color = Metric)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(
    values = c("Unemployment Rate" = "#c0392b",
               "Labor Force Participation Rate" = "#2980b9"),
    name = NULL
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  facet_wrap(~Metric, scales = "free_y", ncol = 1) +
  labs(
    title    = "Unemployment Rate vs. Labor Force Participation Rate",
    subtitle = "LFPR declined steadily from 2008–2015, even as unemployment improved — a hidden labor market story",
    x = NULL, y = NULL,
    caption = "Source: Employment, Unemployment & Labor Force Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    legend.position = "none",
    strip.text = element_text(face="bold", size=11),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

---

# Seasonal Patterns

## Does Unemployment Follow a Seasonal Rhythm?

Labor markets have natural seasonal cycles — construction slows in winter, retail spikes during holidays. The chart below shows how unemployment rates behave across months, split by economic era.

```{r seasonal}
seasonal <- df %>%
  group_by(Month, Era) %>%
  summarise(Avg_Rate = mean(`Unemployment Rate`), .groups="drop") %>%
  mutate(MonthLabel = factor(month.abb[Month], levels = month.abb))

ggplot(seasonal, aes(x = MonthLabel, y = Avg_Rate,
                     group = Era, color = Era)) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Pre-Crisis"     = "#27ae60",
    "Crisis"         = "#c0392b",
    "Early Recovery" = "#e67e22",
    "Late Recovery"  = "#2980b9",
    "Expansion"      = "#8e44ad"
  )) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Average Monthly Unemployment Rate by Economic Era",
    subtitle = "Each line represents the average for that month across years in each era",
    x = NULL, y = "Avg Unemployment Rate (%)",
    color = "Economic Era",
    caption = "Source: Employment, Unemployment & Labor Force Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

---

# Employment vs. Unemployed Persons

## The Absolute Numbers Tell a Different Story

While rates are useful, the raw numbers reveal scale. At the crisis peak, over **236,000 people** were unemployed. The number of employed persons, meanwhile, showed strong and consistent growth from 2010 onward.

```{r employed-vs-unemp}
df %>%
  select(Date, Employed, Unemployed) %>%
  pivot_longer(-Date, names_to = "Status", values_to = "Count") %>%
  ggplot(aes(x = Date, y = Count, fill = Status)) +
  geom_area(alpha = 0.75, position = "identity") +
  scale_fill_manual(values = c("Employed" = "#2980b9", "Unemployed" = "#e74c3c"),
                    name = NULL) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Employed vs. Unemployed Persons Over Time",
    subtitle = "Employed (blue, left scale) shows steady post-recession growth; unemployed (red) peaked in 2010",
    x = NULL, y = "Number of Persons",
    caption = "Source: Employment, Unemployment & Labor Force Data"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

---

# Forecasting: Where Was Unemployment Headed?

## Time Series Forecast Using ARIMA

Using the historical monthly unemployment rate data, we fit an **ARIMA model** — a standard time series forecasting approach — to project where the unemployment rate was trending after August 2019.

This is especially interesting knowing what we know now: the COVID-19 pandemic hit in 2020, causing an unprecedented spike. The forecast below shows the *structural trajectory* the labor market was on before that external shock.

```{r forecast, fig.height=7}
# Build time series
unemp_ts <- ts(df$`Unemployment Rate`,
               start = c(2007, 1),
               frequency = 12)

# Fit ARIMA — auto-select best model
fit <- auto.arima(unemp_ts, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)

# Forecast 24 months ahead (to end of 2021)
fc <- forecast(fit, h = 24, level = c(80, 95))

autoplot(fc) +
  geom_hline(yintercept = mean(df$`Unemployment Rate`[df$Year <= 2008]),
             linetype = "dashed", color = "#27ae60", linewidth = 0.8) +
  annotate("text", x = 2008.5, y = mean(df$`Unemployment Rate`[df$Year <= 2008]) + 0.12,
           label = "Pre-crisis avg: ~3.5%", color = "#27ae60", size = 3.5) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title    = "ARIMA Forecast: Unemployment Rate Through 2021",
    subtitle = paste0("Model: ", as.character(fit),
                      " | Forecast shows continued gradual decline absent external shocks"),
    x = NULL, y = "Unemployment Rate (%)",
    caption = "Source: Employment, Unemployment & Labor Force Data | Forecast via auto.arima()"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face="bold", size=15),
    plot.subtitle = element_text(color="#555", size=11),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(color="#999", size=9)
  )
```

```{r arima-summary}
cat("ARIMA Model Summary\n")
cat("===================\n")
summary(fit)
```

## What the Model Says

```{r forecast-table}
fc_df <- data.frame(
  Month = seq(as.Date("2019-09-01"), by = "month", length.out = 24),
  Forecast = round(as.numeric(fc$mean), 2),
  Lower_80  = round(as.numeric(fc$lower[,1]), 2),
  Upper_80  = round(as.numeric(fc$upper[,1]), 2)
)

fc_df %>%
  head(12) %>%
  kable(caption = "12-Month Unemployment Rate Forecast (Sep 2019 – Aug 2020)",
        col.names = c("Month", "Point Forecast (%)", "Lower 80% CI", "Upper 80% CI")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold=TRUE, background="#2980b9", color="white")
```

---

# Year-over-Year Summary Table

```{r yoy-table}
yearly %>%
  rename(
    `Avg Unemployment (%)` = Avg_UnempRate,
    `Avg LFP Rate (%)`     = Avg_LFPRate,
    `Avg Employed`         = Avg_Employed,
    `Avg Unemployed`       = Avg_Unemployed
  ) %>%
  mutate(
    `Avg Employed`   = format(`Avg Employed`,   big.mark=","),
    `Avg Unemployed` = format(`Avg Unemployed`, big.mark=",")
  ) %>%
  kable(caption = "Annual Labor Market Summary — 2007 to 2019") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold=TRUE, background="#2980b9", color="white") %>%
  row_spec(which(yearly$Year %in% c(2009,2010)), background="#fff3cd")
```

---

# Key Takeaways

<div class="takeaway-box">

1. **The 2008 Financial Crisis doubled unemployment in 2 years.** The rate surged from 3.3% in late 2007 to 7.8% by January 2010 — a nearly 150% increase in unemployment.

2. **Recovery was real but slow.** It took until 2015–2016 for unemployment to fall back to pre-crisis levels, representing one of the slowest labor market recoveries in post-war history.

3. **Labor Force Participation tells a darker story.** The LFPR dropped from ~69% in 2008 to under 67% by 2015. This means part of the unemployment improvement came from people leaving the workforce, not from job creation alone.

4. **Employment grew consistently from 2010 onward.** Despite the slow rate recovery, absolute employment numbers rose every year after 2009 — from ~2.8M to ~3.1M employed persons.

5. **The ARIMA forecast projected continued decline.** Based purely on historical trend, unemployment was on track to approach or breach the 3.3% pre-crisis low by 2020–2021 — before COVID-19 changed everything.

</div>

---

*Data sourced from Employment, Unemployment & Labor Force Data. Analysis conducted in R using `readxl`, `dplyr`, `ggplot2`, `forecast`, and `kableExtra`.*
